---
title: "p8105_hw3_sc5154"
author: "Shaohan Chen"
date: "2022-10-11"
output: github_document
---

This is the solution of P8105 Data Science Homework3.

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, message = FALSE)
```

Load necessary packages and the dataset we are about to use.

```{r}
library(p8105.datasets)
library(tidyverse)
library(ggridges)
library(patchwork)
```


## Problem 1
```{r}
data("instacart")

instacart = 
  instacart %>% 
  as_tibble(instacart)
```


```{r}
ins_df = instacart %>%
  janitor::clean_names()
```

The size of the dataset is: there are `r nrow(ins_df)` rows and `r ncol(ins_df)` columns in the dataset.

The variables in the dataset are: `r colnames(ins_df)`.

Key variables include indicators for user, order, and product. Order-level variables include 'order_dow' and  'order_hour_of_day' which describe the day and time of order. Item-level variables include 'product_name', 'department' and 'aisle' which describe specific information of the product. There are also some other descriptive variables like 'add_to_cart_order' which means order in which each product was added to cart, and 'reordered' indicates if this product has been ordered by this user.

In total, there are `r ins_df %>% select(product_id) %>% distinct %>% count` distinct products found in `r ins_df %>% select(user_id, order_id) %>% distinct %>% count` distinct orders from `r ins_df %>% select(user_id) %>% distinct %>% count` distinct users.

Example observations of this dataset can be seen as follows:
```{r}
head(ins_df[1:9],5)
head(ins_df[10:15],5)
```

As you can see, the variable `order_id` takes integer values like 1, `product_name` are characters like 'Bulgarian Yogurt'.

Next we answer the required questions.

*Question 1.* How many aisles are there, and which aisles are the most items ordered from?

* How many aisles are there:
```{r}
aisle_cnt = 
  ins_df %>% 
  select(aisle_id, aisle) %>% 
  distinct %>%
  nrow
```
There are `r aisle_cnt` aisles totally.

* Which aisles are the most items ordered from:
```{r}
most_aisle = 
  ins_df %>%
  group_by(aisle_id, aisle) %>%
  summarize(n_obs = n()) %>%
  arrange(-n_obs, aisle) %>%
  head(5)
most_aisle
```

It can be seen that the top 5 aisles that most items ordered from are:

'fresh vegetables', '	fresh fruits', 'packaged vegetables fruits', 'yogurt', 'packaged cheese'. And 'fresh vegetables' is the most items ordered from.

*Question 2.* Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.


The plot which display aisles with top most number of items ordered:
```{r}
ins_df %>%
  group_by(aisle_id, aisle) %>%
  summarize(n_obs = n()) %>%
  filter(n_obs > 10000) %>%
  arrange(-n_obs, aisle) %>%
  mutate(aisle = factor(aisle)) %>%
  ggplot(aes(x = reorder(aisle, -n_obs), y = n_obs)) +
    geom_col() +
    scale_x_discrete(guide = guide_axis(angle = 45)) +
    labs(x = "Aisle", y = "Number of Items") 
```

*Question 3.* Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.

```{r}
ins_df %>%
  filter(aisle == 'baking ingredients' | aisle == 'dog food care' | aisle == 'packaged vegetables fruits') %>%
  group_by(aisle, product_name) %>%
  summarize(item_order_times = n()) %>%
  mutate(item_rank = min_rank(-item_order_times)) %>%
  arrange(-item_order_times) %>%
  filter(item_rank <= 3) %>%
  knitr::kable()
```

*Question 4.* Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

```{r}
ins_df %>%
  filter(product_name == 'Pink Lady Apples' | product_name == 'Coffee Ice Cream') %>%
  group_by(order_dow, product_name) %>%
  summarize(mean_order_hour = mean(order_hour_of_day)) %>%
  pivot_wider(
    names_from = order_dow,
    values_from = mean_order_hour
  ) %>%
  knitr::kable(digits = 2)
```


## Problem 2

*Question 1.* Load, tidy, and otherwise wrangle the data. Your final dataset should include all originally observed variables and values; have useful variable names; include a weekday vs weekend variable; and encode data with reasonable variable classes. Describe the resulting dataset (e.g. what variables exist, how many observations, etc).

```{r}
accel_df = read_csv("Data/accel_data.csv") %>%
  janitor::clean_names() %>%
  pivot_longer(
    activity_1:activity_1440,
    names_to = 'activity_index',
    names_prefix = 'activity_',
    values_to = 'activity_cnt'
  ) %>%
  mutate(
  day = factor(day, levels = c('Sunday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday')),
  is_weekday = ifelse((day == 'Saturday' | day == 'Sunday'), 0, 1),
  activity_index = as.numeric(activity_index)
  ) %>%
  arrange(week, day_id)
head(accel_df)
```

The resulting dataset has `r nrow(accel_df)` rows and `r ncol(accel_df)` columns.

The existing variables include `r colnames(accel_df)`. Where 'is_weekday' is the new variable to indicate whether it is a weekday or weekend; 'activity_index' and 'activity_cnt' are manually added variables transformed from original observations using `pivot_longer`. The rest of the variables are time for activity record.

*Question 2.* Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each day, and create a table showing these totals. Are any trends apparent?

```{r}
accel_df %>%
  group_by(week, day) %>%
  summarize(daily_activity_tot = sum(activity_cnt)) %>%
  arrange(week, day) %>%
  pivot_wider(
    names_from = day,
    values_from = daily_activity_tot
  ) %>%
  knitr::kable()
```

The daily total activities tend to change with fluctuation within each week. The daily total activity counts on Saturday for last two weeks are both 1440 which means the activity counts is 1 for every minute. That seems to be abnormal and the data on these two days could be considered as outliers.

*Question 3.* Accelerometer data allows the inspection activity over the course of the day. Make a single-panel plot that shows the 24-hour activity time courses for each day and use color to indicate day of the week. Describe in words any patterns or conclusions you can make based on this graph.

```{r}
accel_df %>%
  mutate(
    daily_time = floor((activity_index)/60) +
      (1/60) * (activity_index) %% 60
    ) %>%
  ggplot(aes(x = daily_time, y = activity_cnt, color = day)) +
    geom_line(alpha = 0.3) + 
    theme(legend.position = 'bottom') + 
    scale_x_continuous(limits = c(0,24), breaks = seq(0, 24, 2)) +
    labs(x = 'Hours from Midnight', 
         y = 'Activity Counts', 
         color = 'Day of Week')
```

Based on this graph, we have:

* Generally in each day, the activity counts tend to be lower at late night, midnight and early morning from around 24 pm to 6 am, and higher between 6am to 22 pm. Generally the activity counts would reach a peak between around 11 am to 13 pm, and reach a higher peak between around 20 pm to 22pm.

* There are some extreme points, such as that on Sunday and Friday, when the level becomes extremely high on Sunday noon and Friday night.

## Problem 3

```{r}
data("ny_noaa")
```

For the raw dataset, the dataset has `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns. Variables include `r colnames(ny_noaa)`.

Next ,let's do some data exploration of the dataset.

*Question 1.* Do some data cleaning. Create separate variables for year, month, and day. Ensure observations for temperature, precipitation, and snowfall are given in reasonable units. For snowfall, what are the most commonly observed values? Why?

Clean the datasets.
```{r}
noaa_df = 
  ny_noaa %>%
  janitor::clean_names() %>%
  separate(date, c("year", "month", "day"), sep = "-", convert = TRUE) %>%
  mutate(
    month = month.name[as.integer(month)],
    prcp = prcp / 10,
    tmax = as.numeric(tmax) / 10, 
    tmin = as.numeric(tmin) / 10
    )
```

The resulting dataset has `r nrow(noaa_df)` rows and `r ncol(noaa_df)` columns.

The existing variables include `r colnames(noaa_df)`. Where 'id' is the id of recording station; 'year', 'month', 'station' is the time variable; 'prcp', 'snow', 'snwd', 'tmax' and 'tmin' are weather variables respectively stands for  Precipitation (tenths of mm),  Snowfall (mm), Snow depth (mm), Maximum temperature (tenths of degrees C), Minimum temperature (tenths of degrees C).

```{r}
most_snow = 
  noaa_df %>%
  group_by(snow) %>%
  summarize(n_obs = n()) %>%
  arrange(-n_obs) %>%
  head(1)
```

For snowfall, the most commonly observed values are: `r most_snow$snow` with `r most_snow$n_obs` observations. That's probably because the New York State does not often snow.

*Question 2.* Make a two-panel plot showing the average max temperature in January and in July in each station across years. Is there any observable / interpretable structure? Any outliers?

```{r}
noaa_df %>%
  filter(month == 'January' | month == 'July') %>%
  group_by(month, id, year) %>%
  summarize(mean_tmax = mean(tmax, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = mean_tmax, color = id)) +
  geom_point(size = 0.2, alpha = 0.4) +
  geom_line(alpha = 0.4) +
  theme(legend.position = "none") + 
  labs(x = 'Year', y = "Mean Max Temperature") +
  facet_grid(. ~ month)
```

Based on the two-panel plot, the range of max temperature(tmax) in January is around -10 to 10 (degree C), in July is around 20 to 30. Generally the mean tmax in July is much higher than January. And the temperature difference range in January is around 20, which is larger than in July that is around 10. There is some outliers in January and July. In January there are extremely low tmax lower than -10 and extremely high tmax close to 10. In July there are also extremely high tmax close to 35 and extremely low tmax lower than 15.

In addition, generally the distributions of max temperature in January and July are changing with fluctuation during the past years, with average tmax around 0 in January and average tmax around 27 in July.

*Question 3.* Make a two-panel plot showing (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option); and (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.

```{r}
tmax_vs_tmin = 
  noaa_df %>%
  ggplot(aes(x = tmax, y = tmin)) +
  geom_hex() + 
  labs(x = 'Max Temperature', y = "Min Temperature")

snowfall_dis = 
  noaa_df %>%
  filter(snow > 0 & snow < 100) %>%
  mutate(year = factor(year)) %>%
  ggplot(aes(x = snow, y = year)) + 
  geom_density_ridges(scale = 0.8) +
  labs(x = 'Snowfall', y = "Year")

tmax_vs_tmin + snowfall_dis
```

